Windowed least-squares model reduction for dynamical systems

### 2019-10-06 main.tex
latest, working on notation

\address[a]{Sandia National Laboratories,  Livermore, CA}
\begin{abstract}
This work proposes a \methodNameLower\ (\methodAcronym) model-reduction
	formulation for dynamical systems. The proposed approach sequentially minimizes the
	time-continuous full-order-model residual over a low-dimensional space--time trial
	subspace defined over an arbitrarily defined time window. The approach comprises a generalization 
of existing model reduction approaches, as particular instances of
  the methodology associate with Galerkin,
	least-squares Petrov--Galerkin (LSPG), and space–time LSPG projection. In
	addition, the approach
	addresses key deficiencies in existing model-reduction
	techniques, e.g., the dependence of (space--time) LSPG projection on the
	time discretization and the exponential growth in time exhibited by \textit{a posteriori}
	error bounds for Galerkin and LSPG projection.  We consider two types of
	space--time trial
	subspaces within the proposed framework: one that reduces only the spatial
	dimension of the full-order model, and one that reduces both the spatial and
	temporal dimensions of the full-order model.  For each type of trial
	subspace, we consider two different solution techniques: direct (i.e.,
	discretize-then-optimize) and indirect (i.e., optimize-then-discretize).
	Numerical experiments conducted using the trial subspaces characterized by spatial
	dimension reduction demonstrate that the \methodAcronym\
	approach can yield more accurate and physically relevant solutions \KTC{What
	does `physically relevant' mean? This is imprecise.} than
	Galerkin and LSPG projection.

@@

The paper proceeds as follows: Section~\ref{sec:math}
outlines the mathematical setting for the full-order model, along with the Galerkin, LSPG, and
space--time LSPG projection. Section~\ref{sec:tclspg} outlines the proposed \methodAcronym\
framework. Section~\ref{sec:numerical_techniques}
outlines numerical techniques for solving \methodAcronymROMs, including both direct and
indirect methods. Section~\ref{sec:analysis} provides
equivalence conditions and error analysis for \methodAcronymROMs.
Section~\ref{sec:numerical_experiments} presents numerical experiments.
Section~\ref{sec:conclude} provides conclusions and perspectives.
We denote vector-valued functions with italicized bold symbols (e.g., $\boldsymbol
x$), vectors with standard bold symbols (e.g., $\mathbf{x}$), 
matrices with capital bold symbols (e.g., $\mathbf{X} \equiv \begin{bmatrix}
\mathbf{x}_1 & \cdots & \mathbf{x}_r\end{bmatrix}$), and spaces with
calligraphic symbols (e.g., $\mathcal{X}$).
%Model reduction of nonlinear dynamical systems is a growing research area.
%The least-squares Petrov-Galerkin (LSPG) approach has gained popularity as a model reduction technique for nonlinear dynamical systems. Derived at the fully discrete level, the LSPG approach solves a least-squares problem that minimizes the fully discrete residual of the reduced-order model at each time-step. LSPG has been shown to be an effective formulation for nonlinear model reduction on complex problems of interest.
%\begin{comment} 
\section{Mathematical formulation}\label{sec:math}
	\KTC{HERE}
	We begin by providing the formulation for the full-order model,
	followed by a description of standard model-reduction methods
	classified according to the type of trial subspace they employ.
	\subsection{Full-order model}\label{sec;FOM}
We consider the system of nonlinear ordinary differential equations (ODEs),
\begin{equation}\label{eq:FOM}
 \stateFOMDotArg{t}  = \velocity(\stateFOMArg{}{t},t), \qquad \stateFOM(0) =
	\stateFOMIC,\qquad t \in [0,T]
\end{equation}
where $\stateFOMDot \equiv \frac{d \stateFOM}{dt}$, 
$T \in \mathbb{R}^+$ denotes the final time, $\stateFOM \in \RR{\fomdim}
	\otimes \timeSpace $ with $\stateFOM: [0,T] \rightarrow  \RR{\fomdim}$
	denotes the state implicitly defined as the solution to initial value
	problem~\eqref{eq:FOM}, 
	$\timeSpace$ denotes the set of real-valued functions acting on the time
	domain, 
 $\stateFOMIC \in \mathbb{R}^{\fomdim}$ denotes the initial condition,
	and $\velocity: \mathbb{R}^{\fomdim} \times [0,T] \rightarrow
	\mathbb{R}^{\fomdim}$ with $(\stateyDiscrete,t) \mapsto
	\velocity(\stateyDiscrete,t)$ denotes the velocity, which is possibly
	nonlinear in its first argument. We refer to Eq.~\eqref{eq:FOM} 
	as the ``full-order model" (FOM) ODE.

@@

\subsection{\spatialAcronym\ trial subspaces}
\spatialAcronym\ trial subspaces approximate the FOM ODE solution
	$\stateFOM$ with an approximation $\approxstate$ that resides in an
	affine space--time trial subspace of dimension $K\ll\fomdim$, i.e., 
\begin{equation}\label{eq:spatial_subspace}
\begin{split}
& \stspaceS \defeq \trialspace \otimes \timeSpace +
	\stateIntercept\otimes\onesFunction,\\ 
& \trialspace \defeq \text{Range}(\basismat),  \\%\label{eq:spatial_subspace2}  \\
&\basismat \defeq \begin{bmatrix}  \basisvec_1  & \cdots &  \basisvec_K \end{bmatrix}, %\label{eq:spatial_subspace3} 
\end{split}
\end{equation}
where 
%	$\timeSpace$ denotes the set of real-valued functions acting on the time
%	domain
%	$[0,T]$, 
$\onesFunction\in\timeSpace$ is defined as $\onesFunction:[0,T]\rightarrow\{1\}$,
	$\basismat \in \RRStar{\romdim}{\fomdim}$ denotes the spatial trial subspace with
	$\RRStar{\romdim}{\fomdim}$ denoting the compact Stiefel manifold (i.e.,  $
	\RRStar{\romdim}{\fomdim}\defeq
	\{ \mathbf{X} \in \RR{\fomdim
	\times \romdim}\, \big|\, \mathbf{X}^T \mathbf{X} = \mathbf{I} \}$),
	$\trialspace\subseteq\RR{\fomdim}$ denotes the spatial trial subspace, and
	$\stateIntercept \in \mathbb{R}^{\fomdim}$ denotes the reference state that
	is often taken to be the initial condition (i.e., $\stateIntercept = \stateFOMIC$). 
%where $\basisvec_i \in \mathbb{R}^N$ are orthonormal basis vectors and $\stateIntercept \in \mathbb{R}^{\fomdim}$ the ``intercept''.
%We refer to $\basismat \in \RRStar{\fomdim \times \romdim}$, with $\RRStar{p \times q}$ denoting the set of orthogonal $q$ frames in $p$, as the spatial trial basis, and $\trialspace$ as the spatial trial space. 
The basis vectors $\basisvec_i$, $i=1,\ldots,K$ are typically constructed
using state snapshots, e.g., via
proper orthogonal decomposition (POD)~\cite{berkooz_turbulence_pod} or the reduced-basis method~\cite{rb_1,rb_2,rb_3,NgocCuong2005,Rozza2008}. 
At any time instance $t\in[0,T]$, ROMs that employ the  \spatialAcronym\ trial subspace approximate the state as
\begin{equation}\label{eq:affine_trialspace}
\stateFOMArg{}{t} \approx \approxstate(t) = \basisspace \genstateArg{}{t} + \stateIntercept,
\end{equation}
where $\genstate \in \RR{\romdim} \otimes \timeSpace$ denotes the generalized
coordinates. Substituting the approximation~\eqref{eq:affine_trialspace} into the FOM ODE~\eqref{eq:FOM} and performing orthogonal
$\elltwo$-projection of the initial condition onto the trial subspace yields
the overdetermined system of ODEs
\begin{equation}\label{eq:g_truncation}
\basisspace \frac{d \genstate} {dt}(t)= \velocity(\basisspace
\genstateArg{}{t} + \stateIntercept,t ), \qquad \genstate(0) = \genstateIC, \qquad t \in [0,T].
\end{equation}
Because Eq.~\eqref{eq:g_truncation} is overdetermined, a solution may not
exist. Typically, either \textit{Galerkin projection} or \textit{least-squares
Petrov--Galerkin projection} is employed to reduce the number of equations
such that a unique solution exists. We now describe these two methods.

@@

\frac{d\genstateGalerkin}{dt}(t)= \massArgnt{n}^{-1} \basisspace^T \stweightingMatArg{n} \velocity(\basisspace

@@

\frac{d\genstateGalerkin}{dt}(t) = \underset{ \genstateyDiscrete \in \RR{\romdim}

@@

Thus, the computed velocity $\frac{d\genstateGalerkin}{dt}(t)$ minimizes the

@@

where $\integrand: \RR{M} \times \RR{M} \times[a,b] \rightarrow  \RR{}$ (for arbitrary $M$) with $\integrand: (\stateyDiscrete,\stateyDotDiscrete,t) \mapsto \integrand(\stateyDiscrete,\stateyDotDiscrete,t)$.%, $\statez: [a,b] \rightarrow \RR{M}$ and $\statezDot: [a,b] \rightarrow \RR{M}$.
 We now introduce the function $\statez: [a,b] \rightarrow \RR{M}$ that is implicitly defined as an extremum of~\eqref{eq:el_gen} subject to the boundary condition $\statez(a) = \statez_a$. %To this end, we start by supposing that $\statez$ is a stationary point of $\mathcal{J}$ that satisfies the boundary conditions. 

@@

&: \RR{} \times [a,b] \rightarrow \RR{M}.

>>>>

\address[a]{Sandia National Laboratories,  Livermore, CA}
\begin{abstract}
This work proposes a \methodNameLower\ (\methodAcronym) model-reduction
	formulation for dynamical systems. The proposed approach sequentially minimizes the
	time-continuous full-order-model residual over a low-dimensional space--time trial
	subspace defined over an arbitrarily defined time window. The approach comprises a generalization 
of existing model reduction approaches, as particular instances of
  the methodology associate with Galerkin,
	least-squares Petrov--Galerkin (LSPG), and space–time LSPG projection. In
	addition, the approach
	addresses key deficiencies in existing model-reduction
	techniques, e.g., the dependence of (space--time) LSPG projection on the
	time discretization and the exponential growth in time exhibited by \textit{a posteriori}
	error bounds for Galerkin and LSPG projection.  We consider two types of
	space--time trial
	subspaces within the proposed framework: one that reduces only the spatial
	dimension of the full-order model, and one that reduces both the spatial and
	temporal dimensions of the full-order model.  For each type of trial
	subspace, we consider two different solution techniques: direct (i.e.,
	discretize-then-optimize) and indirect (i.e., optimize-then-discretize).
	Numerical experiments conducted using the trial subspaces characterized by spatial
	dimension reduction demonstrate that the \methodAcronym\
	approach can yield more accurate solutions that better satisfy the governing equations than
	Galerkin and LSPG projection.

@@

The paper proceeds as follows: Section~\ref{sec:math}
outlines the mathematical setting for the full-order model, along with the Galerkin, LSPG, and
space--time LSPG projection. Section~\ref{sec:tclspg} outlines the proposed \methodAcronym\
framework. Section~\ref{sec:numerical_techniques}
outlines numerical techniques for solving \methodAcronymROMs, including both direct and
indirect methods. Section~\ref{sec:analysis} provides
equivalence conditions and error analysis for \methodAcronymROMs.
Section~\ref{sec:numerical_experiments} presents numerical experiments.
Section~\ref{sec:conclude} provides conclusions and perspectives.
We denote vector-valued functions with italicized bold symbols (e.g., $\boldsymbol
x$), vectors with standard bold symbols (e.g., $\mathbf{x}$), 
matrices with capital bold symbols (e.g., $\mathbf{X} \equiv \begin{bmatrix}
\mathbf{x}_1 & \cdots & \mathbf{x}_r\end{bmatrix}$), and spaces with
calligraphic symbols (e.g., $\mathcal{X}$). We additionally denote differentiation of a time-dependent 
function with respect to time as $\stateDot \equiv \frac{d \state}{dt}$.
%Model reduction of nonlinear dynamical systems is a growing research area.
%The least-squares Petrov-Galerkin (LSPG) approach has gained popularity as a model reduction technique for nonlinear dynamical systems. Derived at the fully discrete level, the LSPG approach solves a least-squares problem that minimizes the fully discrete residual of the reduced-order model at each time-step. LSPG has been shown to be an effective formulation for nonlinear model reduction on complex problems of interest.
%\begin{comment} 
\section{Mathematical formulation}\label{sec:math}
	\KTC{HERE}
	We begin by providing the formulation for the full-order model,
	followed by a description of standard model-reduction methods
	classified according to the type of trial subspace they employ.
	\subsection{Full-order model}\label{sec;FOM}
We consider the system of nonlinear ordinary differential equations (ODEs),
\begin{equation}\label{eq:FOM}
 \stateFOMDotArg{t}  = \velocity(\stateFOMArg{}{t},t), \qquad \stateFOM(0) =
	\stateFOMIC,\qquad t \in [0,T]
\end{equation}
where $\stateFOMDot \equiv \frac{d \stateFOM}{dt}$, 
$T \in \mathbb{R}^+$ denotes the final time, $\stateFOM \in \RR{\fomdim}
	\otimes \timeSpace $ with $\stateFOM: [0,T] \rightarrow  \RR{\fomdim}$
	denotes the state implicitly defined as the solution to initial value
	problem~\eqref{eq:FOM}, 
	$\timeSpace$ denotes the set of real-valued functions acting on the time
	domain, 
 $\stateFOMIC \in \mathbb{R}^{\fomdim}$ denotes the initial condition,
	and $\velocity: \mathbb{R}^{\fomdim} \times [0,T] \rightarrow
	\mathbb{R}^{\fomdim}$ with $(\stateyDiscrete,\timeDumArg) \mapsto
	\velocity(\stateyDiscrete,\timeDumArg)$ denotes the velocity, which is possibly
	nonlinear in its first argument. We refer to Eq.~\eqref{eq:FOM} 
	as the ``full-order model" (FOM) ODE.

@@

\subsection{\spatialAcronym\ trial subspaces}
\spatialAcronym\ trial subspaces approximate the FOM ODE solution
	$\stateFOM$ with an approximation $\approxstate$ that resides in an
	affine space--time trial subspace of dimension $K\ll\fomdim$, i.e., 
\begin{equation}\label{eq:spatial_subspace}
\begin{split}
& \stspaceS \defeq \trialspace \otimes \timeSpace +
	\stateIntercept\otimes\onesFunction,\\ 
& \trialspace \defeq \text{Range}(\basismat),  \\%\label{eq:spatial_subspace2}  \\
&\basismat \defeq \begin{bmatrix}  \basisvec_1  & \cdots &  \basisvec_K \end{bmatrix}, %\label{eq:spatial_subspace3} 
\end{split}
\end{equation}
where 
%	$\timeSpace$ denotes the set of real-valued functions acting on the time
%	domain
%	$[0,T]$, 
$\onesFunction\in\timeSpace$ is defined as $\onesFunction:[0,T]\rightarrow\{1\}$,
	$\basismat \in \RRStar{\romdim}{\fomdim}$ denotes the spatial trial subspace with
	$\RRStar{\romdim}{\fomdim}$ denoting the compact Stiefel manifold (i.e.,  $
	\RRStar{\romdim}{\fomdim}\defeq
	\{ \mathbf{X} \in \RR{\fomdim
	\times \romdim}\, \big|\, \mathbf{X}^T \mathbf{X} = \mathbf{I} \}$),
	$\trialspace\subseteq\RR{\fomdim}$ denotes the spatial trial subspace, and
	$\stateIntercept \in \mathbb{R}^{\fomdim}$ denotes the reference state that
	is often taken to be the initial condition (i.e., $\stateIntercept = \stateFOMIC$). 
%where $\basisvec_i \in \mathbb{R}^N$ are orthonormal basis vectors and $\stateIntercept \in \mathbb{R}^{\fomdim}$ the ``intercept''.
%We refer to $\basismat \in \RRStar{\fomdim \times \romdim}$, with $\RRStar{p \times q}$ denoting the set of orthogonal $q$ frames in $p$, as the spatial trial basis, and $\trialspace$ as the spatial trial space. 
The basis vectors $\basisvec_i$, $i=1,\ldots,K$ are typically constructed
using state snapshots, e.g., via
proper orthogonal decomposition (POD)~\cite{berkooz_turbulence_pod} or the reduced-basis method~\cite{rb_1,rb_2,rb_3,NgocCuong2005,Rozza2008}. 
At any time instance $t\in[0,T]$, ROMs that employ the  \spatialAcronym\ trial subspace approximate the state as
\begin{equation}\label{eq:affine_trialspace}
\stateFOMArg{}{t} \approx \approxstate(t) = \basisspace \genstateArg{}{t} + \stateIntercept,
\end{equation}
where $\genstate \in \RR{\romdim} \otimes \timeSpace$ denotes the generalized
coordinates. Substituting the approximation~\eqref{eq:affine_trialspace} into the FOM ODE~\eqref{eq:FOM} and performing orthogonal
$\elltwo$-projection of the initial condition onto the trial subspace yields
the overdetermined system of ODEs
\begin{equation}\label{eq:g_truncation}
\basisspace \genstateDotArg{}{t} = \velocity(\basisspace
\genstateArg{}{t} + \stateIntercept,t ), \qquad \genstate(0) = \genstateIC, \qquad t \in [0,T].
\end{equation}
Because Eq.~\eqref{eq:g_truncation} is overdetermined, a solution may not
exist. Typically, either \textit{Galerkin projection} or \textit{least-squares
Petrov--Galerkin projection} is employed to reduce the number of equations
such that a unique solution exists. We now describe these two methods.

@@

\genstateGalerkinDotArg{}{t} = \massArgnt{n}^{-1} \basisspace^T \stweightingMatArg{n} \velocity(\basisspace

@@

\genstateGalerkinDotArg{}{t} = \underset{ \genstateyDiscrete \in \RR{\romdim}

@@

Thus, the computed velocity $\genstateGalerkinDotArg{}{t}$ minimizes the

@@

where $\integrand: \RR{M} \times \RR{M} \times[a,b] \rightarrow  \RR{}$ (for arbitrary $M$) with $\integrand: (\stateyDiscrete,\stateyDotDiscrete,t) \mapsto \integrand(\stateyDiscrete,\stateyDotDiscrete,t)$. We now introduce the function $\statez: [a,b] \rightarrow \RR{M}$ that is implicitly defined as an extremum of~\eqref{eq:el_gen} subject to the boundary condition $\statez(a) = \statez_a$. 

@@

&: [a,b] \times \RR{} \rightarrow \RR{M}.
